# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

First, let's load in a few libraries and read in the data.

```{r echo=TRUE}
if (!require(ggplot2)) {
  install.packages("ggplot2")
  library(ggplot2)
}

if (!require(plyr)) {
  install.packages("plyr")
  library(plyr)
}

if (!require(scales)) {
  install.packages("scales")
  library(scales)
}

data <- read.csv(unz("activity.zip", "activity.csv"))
```

## What is mean total number of steps taken per day?

Let's plot the total number of steps per day:  
```{r echo=TRUE}
steps <- ddply(data, "date", summarise, steps = sum(steps, na.rm=TRUE))

ggplot(steps, aes(x = date,y = steps)) + geom_bar(stat = "identity", color="white") + scale_x_discrete(breaks = factor(steps$date)[seq(1, length(steps$date), by = 8)]) + scale_y_continuous(labels = comma) + ggtitle("Steps by Day") + theme(plot.title = element_text(size = rel(2), face="bold", vjust = 1.5))
```

Let's find the mean and median per day:  
```{r echo=TRUE}
stepMean <- mean(steps$steps)
stepMedian <- median(steps$steps)
```

The mean number of steps per day is `r stepMean`.
The median number of steps per day is `r stepMedian`.

## What is the average daily activity pattern?

Let's plot the average number of steps per interval:  
```{r echo=TRUE}
intervals <- ddply(data, "interval", summarise, steps = mean(steps, na.rm=TRUE))

ggplot(intervals, aes(x = interval,y = steps)) + geom_line(stat = "identity") + scale_x_discrete(breaks = factor(seq(100, max(intervals$interval), by = 200))) + scale_y_continuous(labels = comma) + ggtitle("Steps by Interval") + theme(plot.title = element_text(size = rel(2), face="bold", vjust = 1.5))

maxInt <- intervals[intervals$steps == max(intervals$steps),"interval"]
```

The interval with the maximum average number of daily steps is interval `r maxInt`.

## Imputing missing values
```{r echo=TRUE}
numMissing <- sum(is.na(data$steps))
```

There are `r numMissing` missing values in the dataset.

We will impute missing values with the average number of steps *for that interval* across all days, and then plot the resulting histogram.

```{r echo=TRUE}
data2 <- merge(data,rename(intervals, c("steps"="steps2")))
stepsNA <- which(is.na(data2$steps))
data2$steps[stepsNA] <- data2$steps2[stepsNA]

steps2 <- ddply(data2, "date", summarise, steps = sum(steps, na.rm=TRUE))

ggplot(steps2, aes(x = date,y = steps)) + geom_bar(stat = "identity", color="white") + scale_x_discrete(breaks = factor(steps2$date)[seq(1, length(steps2$date), by = 8)]) + scale_y_continuous(labels = comma) + ggtitle("Steps by Day with Imputation") + theme(plot.title = element_text(size = rel(2), face="bold", vjust = 1.5))
```

Let's find the mean and median per day this time:  
```{r echo=TRUE}
stepMean2 <- mean(steps2$steps)
stepMedian2 <- median(steps2$steps)
```

The new mean number of steps per day is `r stepMean2`.  
The new median number of steps per day is `r stepMedian2`.  
Both of these values *increased* by imputing missing values with the average number of steps for that interval. These values are equal because a number of days consisted of nothing but missing values, which were then set to the mean of the other days' intervals.  
Unsurprisingly, there are *more steps per day* after imputing missing values.  

## Are there differences in activity patterns between weekdays and weekends?

Let's plot the difference: 
```{r echo=TRUE}
data2$weekdays <- factor(sapply(weekdays(as.Date(data2$date)), switch, "Saturday" = "Weekend", "Sunday" = "Weekend", "Weekday"))

intervals2 <- ddply(data2, .(interval, weekdays), summarise, steps = mean(steps, na.rm=TRUE))

ggplot(intervals2, aes(x = interval,y = steps)) + geom_line(stat = "identity") + scale_x_continuous(labels = comma) + ggtitle("Steps by Interval, Weekend v. Weekday") + theme(plot.title = element_text(size = rel(2), face="bold", vjust = 1.5)) + facet_grid(weekdays ~ .)
```


